---
format:
  pdf:
    engine: knitr
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \pagenumbering{gobble}
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Atlantic croaker (*`Micropogonias undulates`*)

## Summary

::::: {layout-ncol="2"}
<div>

**Scores**

Vulnerability =

Climate Exposure =

Biological Sensitivity = Moderate

</div>

<div>

**Data Quality**

Model Confidence Score =

Life History Data Quality = 73%

</div>
:::::

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_exposure_summary_table.png")
```

```{r,echo=FALSE, fig.align='left'}

here::here("test_images/Atlantic croaker_sensitivity_summary_table.png")

```

## Changes from Previous Assessment(s)

| Metric               | Current Assessment | Previous Assessment(s) | Change   |
|------------------|-----------------|--------------------|-----------------|
| *Vulnerability*      |                    |                        |          |
| *Exposure*           |                    |                        |          |
| *Sensitivity*        | Moderate           | Low                    | Increase |
| *Directional Effect* | Neutral            | Positive               | Increase |

## Score Drivers

### Vulnerability:

add text here

### Climate Exposure:

add text here

### Biological Sensitivity:

Atlantic croaker received a sensitivity score of **Moderate**. Certainty was high (90%). Three attributes received scores greater than 2.5: 1) Egg and Larval Survival and Settlement Requirements (2.9), 2) Other Stressors (2.7), and 3) Stock Size (2.5).

Croaker eggs and larvae rely on currents and tidal activity for transport to suitable nursery habitats [@sempsrott2017biological]. Larvae are selective feeders, only feeding on a subset of plankton species [@govoni1986selective]. Juveniles are also dependent on native salt marsh habitats, which themselves are highly vulnerable [@guyant2022northeast] and are susceptible to eutrophication and hypoxic events. Croaker had a low sensitivity score in the previous assessment [@hare2016vulnerability]. Egg and Larval Survival and Settlement Requirements drove the previous sensitivity score.

## Data Quality Metrics

### Life History Data Quality:

73% of sensitivity attribute data quality scores were greater than or equal to 2, indicating good data quality. Four attributes were considered data gaps (mean attribute data quality score \< 2): Complexity in Reproductive Strategy, Egg and Larval Survival and Settlement Requirements, Population Growth Rate, and Stock Size/Status.

### Model Confidence:

add text here

## Additional Metrics

### Distribution Shifts:

add text here

### Directional Effect:

The effect of climate change on croaker is estimated to be **neutral** with 89% certainty in expert scores. Warmer temperatures may decrease young-of-year mortality rates. However, juvenile habitats are highly vulnerable to climate change. In the previous assessment, the effects of climate change were estimated to have a positive impact on croaker [@hare2016vulnerability].

## Climate Effects on Abundance and Distribution

As a sub-tropical species, Atlantic croaker may experience both positive and negative effects on their abundance and distribution as a result of climate change. Interannual variability in croaker abundance may be climate related, with colder winters causing increased mortality in overwintering young-of-the-year [@lankford2001low]. Warm winters, however, may result in increased abundances, or “outbursts” of juveniles in the spring that contribute to expansion of the population’s distribution [@hare2007mechanistic]. @diamond2013simulating modeled Atlantic croaker abundance in the Mid-Atlantic with global climate change projections and found increased croaker abundance with lower inter-annual variability under global climate change. Positive responses in sensitivity analyses stemmed from warmer temperatures, and negative responses were due to rising variability in salinity and increases in offshore transport and sea level rise [@diamond2013simulating]. Common juvenile habitat types (submerged aquatic vegetation, salt marshes, and intertidal habitats) are highly vulnerable to climate change [@guyant2022northeast].

## Life History Synopsis

Atlantic croaker are a demersal species found from the Gulf of Maine to Argentina. Along the United States Atlantic coast, they are most abundant from Chesapeake Bay to northern Florida [@asmfc2020addendum]. Atlantic croaker spawn from July - December in inshore shelf and estuarine waters between Delaware Bay and Cape Hatteras, peaking in August - October near Chesapeake Bay [@able1998first]. Their name comes from the croaking noises the fish make during spawning by vibrating their abdominal muscles against their swim bladder [@asmfc2020addendum]. After 30-90 days post-hatch Atlantic croaker larvae are driven inshore by winds [@thorrold1997response] where they migrate to estuarine habitats [@sempsrott2017biological]. Initial fall recruitment of croaker depends on fall continental shelf winds to provide transport into Chesapeake Bay [@norcross1991estuarine]. Recruitment to specific estuaries is influenced by tidal activity and currents, such that eggs and larvae are dependent on transport to suitable habitats for survival [@sempsrott2017biological]. Atlantic croaker larvae are selective feeders, as they eat fewer taxa of suitable size than are present in ambient plankton [@govoni1986selective]. Juvenile croaker are highly dependent on marine, estuarine, and riverine mud and sand substrates, as well as native salt marsh habitats [@guyant2022northeast]. The major prey of young-of-the-year Atlantic croaker are polychaetes, copepods, and mysids. Detritus is also a major component of the juvenile diet [@ffwcc2010croaker] [@sink2011species]. Adult Atlantic croaker perform seasonal migrations, moving northward and inshore to estuaries and bays in the spring - fall and southward and offshore to shelf waters in the winter [@asmfc2020addendum]. They reach maturity by 2 years of age and can live up to 17 years, although lifespans are commonly below 10 years [@asmfc2020addendum]. Adult Atlantic croaker are opportunistic feeders and are known to consume polychaetes, anchovies, mysid shrimp, amphipods, fishes and crabs, and detritus in Chesapeake Bay [@nye2011annual]. Atlantic croaker are managed under Amendment 1 to the Atlantic States Marine Fisheries Commission’s Interstate Fishery Management Plan for Atlantic Croaker (2005) and Addenda I, II, and III (2011, 2014, 2020). [@asmfc2020addendum]. The most recent benchmark stock assessment was completed in 2017, where the current stock status could not be determined because the assessment results were sensitive to certain modeling assumptions, but any further depletion in the stock was unlikely [@asmfc2017croaker].

## Figures

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_exposure_allvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_exposure_impvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_distribution_center_of_gravity.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_distribution_area_of_high_probability.png")
```

```{r,echo=FALSE, out.width="200%", fig.align='left'}
here::here("test_images/Atlantic croaker_sdm.png")
```

## References
