---
format:
  pdf:
    engine: knitr
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \pagenumbering{gobble}
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Atlantic croaker (*`Micropogonias undulates`*)

## Summary

::::: {layout-ncol="2"}
<div>

**Scores**

Vulnerability =

Climate Exposure =

Biological Sensitivity = Moderate

</div>

<div>

**Data Quality**

Model Confidence Score =

Life History Data Quality = 73%

</div>
:::::

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_exposure_summary_table.png")
```

```{r,echo=FALSE, fig.align='left'}

here::here("test_images/Atlantic croaker_sensitivity_summary_table.png")

```

## Changes from Previous Assessment(s)

| Metric               | Current Assessment | Previous Assessment(s) | Change   |
|-------------------|-----------------|---------------------|----------------|
| *Vulnerability*      |                    |                        |          |
| *Exposure*           |                    |                        |          |
| *Sensitivity*        | Moderate           | Low                    | Increase |
| *Directional Effect* | Neutral            | Positive               | Increase |

## Score Drivers

### Vulnerability:

add text here

### Climate Exposure:

add text here

### Biological Sensitivity:

Atlantic croaker received a sensitivity score of **Moderate**. Certainty was high (90%). Three attributes received scores greater than 2.5: 1) Egg and Larval Survival and Settlement Requirements (2.9), 2) Other Stressors (2.7), and 3) Stock Size (2.5).

Croaker eggs and larvae rely on currents and tidal activity for transport to suitable nursery habitats (Sempsrott et al 2017). Larvae are selective feeders, only feeding on a subset of plankton species (Govoni et al 1986). Juveniles are also dependent on native salt marsh habitats, which themselves are highly vulnerable (Guyant et al 2022) and are susceptible to eutrophication and hypoxic events. Croaker had a low sensitivity score in the previous assessment (Hare et al 2016). Egg and Larval Survival and Settlement Requirements drove the previous sensitivity score.

## Data Quality Metrics

### Life History Data Quality:

add text here

### Model Confidence:

add text here

## Additional Metrics

### Distribution Shifts:

add text here

### Directional Effect:

add text here

## Climate Effects on Abundance and Distribution:

## Life History Synopsis:

## Figures

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_exposure_allvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_exposure_impvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_distribution_center_of_gravity.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Atlantic croaker_distribution_area_of_high_probability.png")
```

```{r,echo=FALSE, out.width="200%", fig.align='left'}
here::here("test_images/Atlantic croaker_sdm.png")
```

## References
