---
format:
  pdf:
    engine: knitr
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \pagenumbering{gobble}
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Blueline tilefish (*`Caulolatilus microps`*)

## Summary

::::: {layout-ncol="2"}
<div>

**Scores**

Vulnerability =

Climate Exposure =

Biological Sensitivity =

</div>

<div>

**Data Quality**

Model Confidence Score =

Life History Data Quality =

</div>
:::::

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Blueline tilefish_exposure_summary_table.png")
```

```{r,echo=FALSE, fig.align='left'}

here::here("test_images/Blueline tilefish_sensitivity_summary_table.png")

```

## Changes from Previous Assessment(s)

| Metric               | Current Assessment | Previous Assessment(s) | Change |
|----------------------|--------------------|------------------------|--------|
| *Vulnerability*      |                    |                        |        |
| *Exposure*           |                    |                        |        |
| *Sensitivity*        |                    |                        |        |
| *Directional Effect* |                    |                        |        |

## Score Drivers

### Vulnerability:

add text here

### Climate Exposure:

add text here

### Biological Sensitivity:

add text here

## Data Quality Metrics

### Life History Data Quality:

add text here

### Model Confidence:

add text here

## Additional Metrics

### Distribution Shifts:

add text here

### Directional Effect:

add text here

## Climate Effects on Abundance and Distribution:

## Life History Synopsis:

## Figures

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Blueline tilefish_exposure_allvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Blueline tilefish_exposure_impvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Blueline tilefish_distribution_center_of_gravity.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Blueline tilefish_distribution_area_of_high_probability.png")
```

```{r,echo=FALSE, out.width="200%", fig.align='left'}
here::here("test_images/Blueline tilefish_sdm.png")
```

## References
