---
format:
  pdf:
    engine: knitr
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \pagenumbering{gobble}
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Bluefish (*`Pomatomus saltatrix`*)

## Summary

::::: {layout-ncol="2"}
<div>

**Scores**

Vulnerability =

Climate Exposure =

Biological Sensitivity = Low

</div>

<div>

**Data Quality**

Model Confidence Score =

Life History Data Quality = 97%

</div>
:::::

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Bluefish_exposure_summary_table.png")
```

```{r,echo=FALSE, fig.align='left'}

here::here("test_images/Bluefish_sensitivity_summary_table.png")

```

## Changes from Previous Assessment(s)

| Metric               | Current Assessment | Previous Assessment(s) | Change   |
|-------------------|-----------------|---------------------|----------------|
| *Vulnerability*      |                    |                        |          |
| *Exposure*           |                    |                        |          |
| *Sensitivity*        | Low                | Low                    | None     |
| *Directional Effect* | Neutral            | Positive               | Decrease |

## Score Drivers

### Vulnerability:

add text here

### Climate Exposure:

add text here

### Biological Sensitivity:

Bluefish received a **Low** sensitivity score, with a certainty of 69%. The only attribute that received a score greater than 2.5 was Juvenile Habitat Requirements (4). Bluefish rely entirely on salt marshes and submerged aquatic vegetation as juveniles; both habitats are considered highly vulnerable to future climate change [@guyant2022northeast].

Bluefish also received a Low sensitivity score in the 2016 Assessment. Similar to our results, Early Life History Requirements was the only attribute to score above 2.5 [@hare2016vulnerability].

## Data Quality Metrics

### Life History Data Quality:

97% of data quality scores were greater than or equal to 2, indicating excellent data quality. No data gaps (mean attribute data quality score \< 2) were identified.

### Model Confidence:

add text here

## Additional Metrics

### Distribution Shifts:

add text here

### Directional Effect:

Bluefish are expected to be **neutrally** impacted by climate change. Expert certainty was 95%. Distributions are shifting north and east and warming temperatures may expand the spawning season [@Tyrell2022]. However, there is a concern that seasonal mismatches may occur if juveniles arrive in estuaries before their prey, which may impact recruitment [@Taylor2006]. In the original Assessment, bluefish were predicted to be positively impacted by climate change due to expanding distributions and spawning seasons [@hare2016vulnerability].

## Climate Effects on Abundance and Distribution

There has been a documented shift in bluefish distribution, spawning, habitat use in the Mediterranean [@Sabates2012; @VillegasHernandez2015]. Bluefish distributions in the Northwest Atlantic seem to be shifting north and east [@Tyrell2022]. Increasing fall sea surface temperatures may extend spawning season further into the calendar year [@Tyrell2022]. If warming alters spawning phenology, there are concerns that juveniles may arrive in estuaries before their prey. Seasonal mismatches have been shown to impact growth, body condition, and reproductive success [@Taylor2006]. Changing distributions of prey species as a result of climate change may alter interactions between bluefish and their prey [@Whitfield2026]. 

## Life History Synopsis

Bluefish inhabit temperate and tropical surface waters in the North Atlantic from Maine to the Atlantic coast of Florida [@Shepherd2006]. Bluefish in the North Atlantic are a single, migratory stock with two discrete spawning periods. The first occurs during March-May near the edge of the continental shelf of the South-Atlantic Bight. The second occurs between June and August in the Mid-Atlantic Bight [@Shepherd2006]. Eggs hatch after 46-48 hours at 18 - 22.2°C to pelagic larvae [@Deuel1966]. Larvae spawned in the South Atlantic (spring-spawned cohort) are subject to advection north via the Gulf Stream and inshore via warm core rings. Eggs in the southern part of the Mid-Atlantic Bight may be advected south and offshore [@Shepherd2006]. Transport of larvae and/or juveniles across the shelf- slope region may be aided by wind-driven surface flow [@Hare1996]. @Sherman1984 note that wide distribution of bluefish larvae over the shelf reduces their vulnerability to the predator field and environmental perturbation. Larvae prey on copepods [@Shepherd2006]. Peak abundance of larval bluefish in Southern New England and the Mid-Atlantic Bight coincided with zooplankton production [@Sherman1984]. Changes in prey availability and distribution can impact bluefish condition and reproductive success [@Taylor2006]. Juvenile bluefish inhabit estuaries, bays, and coastal oceans with sand, oyster beds, and submerged aquatic vegetation habitats [@McBride1991]. Juveniles enter estuaries when temperatures are greater than 13-15°C and leave when temperatures fall below 10-12°C [@Smith1994]. Juvenile migration into estuarine habitats appears tied to a shift in eating other fish. In estuaries, adults and juveniles eat whatever taxa are locally abundant (fish, crustaceans, shrimp, and polychaetes). In ocean habitats, young-of-the-year bluefish switch to piscivory predation with increasing size, similar to estuarine habitats [@Marks1996]. Bluefish are opportunistic foragers, and the composition of adult bluefish diet depends largely on the available prey, but they consume similar prey items as juvenile bluefish [@Shepherd2006]. Bluefish have long been recognized as a notable predator of forage fish [@Tyrell2022]. Adult bluefish conduct seasonal migrations following warmer waters, overwintering in the south and moving north in the spring and summer [@Shepherd2006]. Migrations may be cued by photoperiod as well, and have been observed to shift northward in warm years [@Nye2009]. Shortfin mako sharks are identified as a major predator of bluefish and may consume more individuals than removed by the fishery [@Wood2009]. Most bluefish are mature by age 2, with approximately half maturing by age 1 [@NEFSC2015_bluefish]. The maximum age for bluefish in the North Atlantic is 12 years [@Salerno2001]. In the Northwest Atlantic, bluefish are managed by the Mid-Atlantic Fishery Management Council and Atlantic States Marine Fisheries Commission. The last stock assessment occurred in 2025, where bluefish were found to not be overfished nor is overfishing occurring [@NEFSC2025_bluefish]. 

## Figures

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Bluefish_exposure_allvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Bluefish_exposure_impvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Bluefish_distribution_center_of_gravity.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Bluefish_distribution_area_of_high_probability.png")
```

```{r,echo=FALSE, out.width="200%", fig.align='left'}
here::here("test_images/Bluefish_sdm.png")
```

## References
