---
format:
  pdf:
    engine: knitr
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \pagenumbering{gobble}
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Eastern oyster (*`Crassostrea virginica`*)

## Summary

::::: {layout-ncol="2"}
<div>

**Scores**

Vulnerability =

Climate Exposure =

Biological Sensitivity = Very High

</div>

<div>

**Data Quality**

Model Confidence Score =

Life History Data Quality = 90%

</div>
:::::

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Eastern oyster_exposure_summary_table.png")
```

```{r,echo=FALSE, fig.align='left'}

here::here("test_images/Eastern oyster_sensitivity_summary_table.png")

```

## Changes from Previous Assessment(s)

| Metric               | Current Assessment | Previous Assessment(s) | Change   |
|-------------------|-----------------|---------------------|----------------|
| *Vulnerability*      |                    |                        |          |
| *Exposure*           |                    |                        |          |
| *Sensitivity*        | Very High          | High                   | Increase |
| *Directional Effect* | Neutral            | Negative               | Increase |

## Score Drivers

### Vulnerability:

add text here

### Climate Exposure:

add text here

### Biological Sensitivity:

The Eastern oyster received a **Very High** sensitivity score, with 100% certainty according to bootstrap analysis. Five attributes had scores greater than 3.5: Adult Mobility (4), Other Stressors (4), Relationship to Calcified Organisms (3.9), Complexity in Reproductive Strategy (3.8), and Egg and Larval Survival and Settlement Requirements (3.5).

Eastern Oysters are sessile once they settle out of the plankton, and settle in estuarine habitats. As a result, they are exposed to a variety of stressors due to urbanization impacts such as hypoxia, loss of seagrass beds, dredging, and changes to freshwater inputs [@guyant2022northeast]. Spawning is initiated by warmer temperatures. Settling out of the plankton also is dependent on temperature and the presence of conspecifics. Because recruitment is density dependent, population collapse can be sudden.

In the 2016 Assessment, Eastern Oysters received a High sensitivity score [@hare2016vulnerability]. This was driven by Sensitivity to Ocean Acidification, Adult Mobility, and Other Stressors, similar to our results.

## Data Quality Metrics

### Life History Data Quality:

90% of data quality scores were greater than or equal to 2, indicating excellent data quality. Population Growth Rate and Stock Size/Status were identified as data gaps (mean attribute data quality score \< 2).

### Model Confidence:

add text here

## Additional Metrics

### Distribution Shifts:

add text here

### Directional Effect:

Experts predicted that climate change would have a **neutral** effect on Eastern Oysters. Certainty was low at 60%. While oysters live in vulnerable estuarine habitats [@guyant2022northeast], spawning occurs throughout the year when temperatures are warm [@berringan1991oyster]. Aquaculture programs also help contribute to population recovery [@mackenzie1981biotic]. In the previous Assessment, experts predicted that climate change would have a negative impact on Eastern Oysters.

## Climate Effects on Abundance and Distribution

Climate change, warmer temperatures, and perhaps different tidal ranges, may increase competition for resources with current competitors, like hooked mussel, invasive species such as *Didemnum* sp., or increase exposure to predators like veined rapa whelk [@white1996predators]. Depletion of oyster beds will also likely have negative effects for many marine species and habitats, for they provide refuge and homes for many fish and invertebrates species and help prevent erosion [@eobrt2007status]. Hurricanes can cover oysters in silt, wash them off beds, or impose extended “freshettes” on oyster beds to the degree that they are significantly impaired and need restoration. This may become more problematic, for hurricanes are thought to increase with increased warming temperatures and climate change [@perret1999oyster]. Since adult oysters are sessile and inhabit mainly highly vulnerable habitats in areas with high climate exposure, their spawning habitat is also highly vulnerable [@farr2021assessment].

## Life History Synopsis

Eastern oysters are found along the eastern U.S. from the Gulf of St. Lawrence to the Gulf of America. They inhabit brackish and salty waters, forming beds or reefs by attaching to each other or other hard substrate [@ozbay2014eastern]. Eastern oysters are a keystone species, and provide valuable habitat for benthic and intertidal organisms. Oysters are sessile filter feeders, contributing to increased water quality through reduced sediment suspension and nutrient uptake and sequestration [@ozbay2014eastern]. Eastern oyster populations in the Mid-Atlantic and Gulf declined in the 19th and 20th centuries, mainly from anthropogenic factors such as land use, habitat degradation, and overharvesting. Disease mortality from Dermo and MSX continues to diminish oyster recovery [@eobrt2007status]. Aquaculture production of oysters contributes to population recovery and habitat restoration programs across the East Coast [@mackenzie1981biotic]. Oysters spawn during the summer in the North and Mid-Atlantic and throughout most of the year in the South Atlantic when temperatures remain above 20°C and salinities above 10ppt [@berringan1991oyster]. The first larval stage (trochophore) is formed four to six hours following fertilization. Larvae normally settle two to three weeks after hatching, but can last up to 2 months [@eobrt2007status]. Larvae move both by passive and active (as development progresses) transport, although larval transport in estuaries seems to be minimal. A balance is reached where the larvae can find suitable areas to settle while still being retained in estuaries [@eobrt2007status]. Settlement ends at the end of the larval stage, and the larval duration is highly temperature dependent [@shumway1996natural]. @fritz1984selective found that planktotrophic oyster larvae fed upon a variety of phytoplankton taxa (coccoids, centrate diatoms, pennate diatoms, dinoflagellates, and flagellates). @baldwin1991omnivorous found that the larvae are actually omnivorous, preying on both autotrophic and heterotrophic plankton. Predation on oyster larvae is extremely high by filter-feeders in the water column and in the benthos. Along the Atlantic coast, drills have been responsible for mortality on 33% of spat in Long Island Sound. Oysters can compete with other mussels and barnacles for shell and substrate spaces, resulting in increased mortality when competitors are abundant [@kennedy1996biology]. Juveniles and adults feed primarily on phytoplankton and suspended detritus, relying on food-laden water being moved past them in order to extract food from the water column [@langdon1996digestion]. Maturity occurs between 1 and 3 years of age [@fsuml_oyster_lifecycle]. Several ages have been proposed, the most common being 20 years. Oysters in Maryland appear to survive at
at least 10 years based on plantings at restoration sites which are not thought to have received natural spat sets. It is likely that the maximum longevity is substantially longer than 10 years [@mddnr2025stock]. Wild and aquaculture harvest is managed by individual jurisdictions or states, and management guidelines vary [@ozbay2014eastern].


## Figures

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Eastern oyster_exposure_allvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Eastern oyster_exposure_impvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Eastern oyster_distribution_center_of_gravity.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Eastern oyster_distribution_area_of_high_probability.png")
```

```{r,echo=FALSE, out.width="200%", fig.align='left'}
here::here("test_images/Eastern oyster_sdm.png")
```

## References
