---
format:
  pdf:
    engine: knitr
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \pagenumbering{gobble}
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Red drum (*`Sciaenops ocellatus`*)

## Summary

::::: {layout-ncol="2"}
<div>

**Scores**

Vulnerability =

Climate Exposure =

Biological Sensitivity = Moderate

</div>

<div>

**Data Quality**

Model Confidence Score =

Life History Data Quality = 95%

</div>
:::::

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Red drum_exposure_summary_table.png")
```

```{r,echo=FALSE, fig.align='left'}

here::here("test_images/Red drum_sensitivity_summary_table.png")

```

## Changes from Previous Assessment(s)

| Metric               | Current Assessment | Previous Assessment(s) | Change   |
|-------------------|-----------------|---------------------|----------------|
| *Vulnerability*      |                    |                        |          |
| *Exposure*           |                    |                        |          |
| *Sensitivity*        | Moderate           | Moderate               | None     |
| *Directional Effect* | Neutral            | Positive               | Increase |

## Score Drivers

### Vulnerability:

add text here

### Climate Exposure:

add text here

### Biological Sensitivity:

Red Drum received a **Moderate** sensitivity score with 50% certainty. This was driven by five attributes with scores over 2.5: Complexity in Reproductive Strategy (3.3), Population Growth Rate (3), Other Stressors (2.9), Egg and Larval Survival and Settlement Requirements (2.6), and Stock Size (2.6).

Red Drum utilize spawning aggregations and spawn in estuaries, which are known vulnerable habitats [@lowerre2016assessing; @guyant2022northeast]. It is hypothesized that red tides impact spawning aggregations in the Gulf of America [@lowerre2016assessing]. Males also use sounds to attract females to spawning locations [@looby2022quantitative; @montie2016long]. They mature at 3-4 years of age, and can live up to 62 years, making them a very long-lived species [@asmfc2024reddrum]. Larvae may rely on currents for transport into nursery habitats, which include seagrass beds.

Seagrass beds may contain an olfactory cue for settlement, and are highly vulnerable habitats in the Northeast US [@asmfc2024reddrum; @havel2016settlement; @comyns1989age; @guyant2022northeast]. Adults are also dependent on shellfish reef habitats and estuarine water column habitats that are highly vulnerable [@guyant2022northeast]. Invasive species may also displace the native plant and prey species that red drum rely on [@asmfc2024reddrum].

While the stock within the Northeast US is not overfished and overfishing is not occurring, models for the northern stock had high uncertainty and instability and could not be used to determine stock status in the most recent assessment [@asmfc2024reddrum].

In the previous Climate Vulnerability Assessment for the Northeast Shelf, Red Drum also received a Moderate sensitivity score. This score was driven by similar attributes to our score: Population Growth Rate, Spawning Cycle, and Early Life History Requirements [@hare2016vulnerability].

## Data Quality Metrics

### Life History Data Quality:

95% of data quality scores were greater than or equal to 2, indicating excellent data quality. No data gaps (mean attribute data quality score \< 2) were identified.

### Model Confidence:

add text here

## Additional Metrics

### Distribution Shifts:

add text here

### Directional Effect:

Experts predict that climate change will have a **neutral** effect on Red Drum. Certainty was 85%. Higher temperatures have been correlated with strong recruitment events [@goldberg2022analysis]. However, Red Drum occupy vulnerable estuarine habitats, and rely on highly vulnerable habitats such as seagrass beds and shellfish reef habitats [@guyant2022northeast]. In the previous Assessment for the Northeast Shelf, experts predicted Red Drum would be positively impacted by climate change, albeit with low certainty [@hare2016vulnerability].

## Climate Effects on Abundance and Distribution

Red drum recruitment has been correlated with wind directions and sea surface temperature [@asmfc2024reddrum]. In North Carolina, the duration and magnitude of N and NE winds in August was twice as strong in high recruitment years. Elevated nearshore sea surface temperatures and coastal productivity (high chlorophyll a) were positively correlated with strong year classes [@goldberg2022analysis]. Red drum acclimated to higher temperatures had a higher standard metabolic rate, decreased critical oxygen threshold, and reduced ability to recover from exercise. Crude oil exposure (hypoxic conditions) also resulted in lower metabolic rates and reduced ability to recover from exercise, and combined with climate-induced increases in temperature, are likely to impact red drum respiratory performance and fitness [@ackerly2021effects]. Laboratory experiments on early life stage red drum resulted in decreased survival and increased heart rate when exposed to elevated pCO^2^ ; however the effects were less severe on red drum compared to similar studies on other species [@lonthair2017early]. 

## Life History Synopsis

Red drum, or redfish, are a popular recreational fish species found in estuarine and inshore waters from Massachusetts to Florida. The stock is divided into northern and southern subunits from New Jersey to North Carolina and South Carolina to Florida, respectively [@asmfc2024reddrum]. Spawning occurs within 45-60 days between August and October and where currents can move larvae/eggs into nearby estuaries and nursery habitats [@asmfc2024reddrum]. Optimal spawning temperatures in laboratory studies ranged between 22 - 25°C, with spawning observed from 22 - 30°C [@asmfc2024reddrum]. Eggs hatch within 24-36 hours [@peters1987early]. Larvae settle at approximately 17 days old in Florida estuaries [@rooker1998vulnerability]. Temperatures of 25°C and salinities around 30 are optimal for hatching success and larval survival. Larvae growth rates were also correlated with ocean temperatures in the Gulf of America [@asmfc2024reddrum; @comyns1989age]. Larvae in Florida consume abundant copepods, whereas juveniles consume abundance amphipods, shrimp, and mysids, switching to crabs and fish as they grow [@peters1987early]. Juveniles rely on structural nursery habitats such as seagrasses for protection from predators. Juveniles are most abundant in estuarine waters and inlets, while fish older than age four inhabit deeper waters [@rooker1998vulnerability]. Adults in the Gulf of America (TX) consume white shrimp, gulf menhaden, and blue crab. Dominant prey species shift seasonally (shrimp in fall, menhaden in spring) while some prey species are consistent throughout the year (crab) [@scharf2000feeding]. Age at maturity in red drum off North Carolina is 3-4 years, with a maximum age of 62 years [@asmfc2024reddrum]. Red drum move out of estuaries into coastal waters as they mature. Seasonal movements with seasonal site fidelity have been observed in Georgia but some populations (particularly in the southern stock) are resident year round. Seasonal migrations between Virginia and North Carolina have been observed between May and September [@asmfc2024reddrum]. The northern stock is managed by the Atlantic States Marine Fisheries Commission. The northern stock is not overfished and overfishing is not occurring, however the southern stock is overfished and overfishing is currently occurring [@asmfc2024reddrum].

## Figures

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Red drum_exposure_allvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Red drum_exposure_impvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Red drum_distribution_center_of_gravity.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Red drum_distribution_area_of_high_probability.png")
```

```{r,echo=FALSE, out.width="200%", fig.align='left'}
here::here("test_images/Red drum_sdm.png")
```

## References
