---
format:
  pdf:
    engine: knitr
    geometry:
      - top=20mm
      - bottom=20mm
      - left=20mm
      - right=20mm
    include-in-header: 
      text: |
        \usepackage{fvextra}
        \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
        \pagenumbering{gobble}
---

```{r setup}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

# Scup (*`Stenotomus chrysops`*)

## Summary

::::: {layout-ncol="2"}
<div>

**Scores**

Vulnerability =

Climate Exposure =

Biological Sensitivity = Moderate

</div>

<div>

**Data Quality**

Model Confidence Score =

Life History Data Quality = 98%

</div>
:::::

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Scup_exposure_summary_table.png")
```

```{r,echo=FALSE, fig.align='left'}

here::here("test_images/Scup_sensitivity_summary_table.png")

```

## Changes from Previous Assessment(s)

| Metric               | Current Assessment | Previous Assessment(s) | Change   |
|-------------------|-----------------|---------------------|----------------|
| *Vulnerability*      |                    |                        |          |
| *Exposure*           |                    |                        |          |
| *Sensitivity*        | Moderate           | Low                    | Increase |
| *Directional Effect* | Neutral            | Positive               | Increase |

## Score Drivers

### Vulnerability:

add text here

### Climate Exposure:

add text here

### Biological Sensitivity:

Scup received a **Moderate** sensitivity score with a certainty of 59%. A total of four attributes had scores greater than 2.5: Spawning Cycle (3), Complexity in Reproductive Strategy (2.9), Dispersal of Early Life Stages (2.9), and Other Stressors (2.8).

Scup only spawn once per year between May and August [@steimle1999_scup]. Spawning occurs in inshore waters once they are warmer than 10C and rely on highly vulnerable habitats such as submerged aquatic vegetation and estuarine habitats [@steimle1999_scup; @guyant2022northeast]. Spawning may also rely on soniferous behaviors [@looby2022quantitative]. Eggs and larvae do not disperse far from spawning locations [@steimle1999_scup]. Adults utilize benthic habitats on the continental shelf and in estuaries. As a result, they are exposed to a variety of stressors due to urbanization impacts such as hypoxia, changes to freshwater inputs, and dredging [@guyant2022northeast].

In the 2016 Climate Vulnerability Assessment, Scup received a Low sensitivity score. Similar to our analysis, this score was driven by Spawning Cycle and Early Life History Requirements, as well as Population Growth Rate [@hare2016vulnerability]. In our analysis, Population Growth Rate received a score of 2.4.

## Data Quality Metrics

### Life History Data Quality:

98% of data quality scores were greater than or equal to 2, indicating excellent data quality. No data gaps (mean attribute data quality score \< 2) were identified.

### Model Confidence:

add text here

## Additional Metrics

### Distribution Shifts:

add text here

### Directional Effect:

Experts predicted that climate change will have a **neutral** effect on Scup. Certainty was 93%. Scup distributions have shifted northward and offshore [@bell2015disentangling; @palacios2023quantifying], likely in search of preferred temperatures, which range between 9-24C [@steimle1999_scup], but their center of distribution has not changed [@roday2026recreational]. Juveniles and adults both utilize vulnerable estuarine habitats [@guyant2022northeast]. In the previous assessment, experts predicted that Scup would be positively impacted by climate change, however, uncertainty was high [@hare2016vulnerability].

## Climate Effects on Abundance and Distribution

Scup distributions in the spring and fall have shifted northward and offshore, which may necessitate new management strategies [@bell2015disentangling; @palacios2023quantifying]. @roday2026recreational analyzed fish distribution shifts using recreational catch and fishery independent data. Scup distributions showed a moderate range expansion but little to no change to center of distribution [@roday2026recreational]. Juvenile scup exposed to elevated pCO^2^ in a laboratory setting had no significant effect on growth, survival, or otolith condition. Scup spawning areas in Long Island Sound are already exposed to high pCO^2^, indicating some tolerance to ocean acidification [@perry2015effect]. 

## Life History Synopsis

Scup are found in the Northwest Atlantic from Cape Cod, Massachusetts to Cape Hatteras, North Carolina [@steimle1999_scup]. There are two U.S. stocks: Mid-Atlantic/New England (Atlantic coast) and South Atlantic [@nefsc2025scup]. Spawning occurs once per year in coastal bays and estuaries in May-August, beginning with lower latitudes [@steimle1999_scup]. Eggs take 2-3 days to hatch depending on temperature. The larval stage lasts from 2 mm TL to 18-19 TL; the duration of this period is unclear. Larval distributions are highest in inshore bays and estuaries, suggesting that eggs and larvae do not travel far during planktonic stages [@steimle1999_scup]. The biggest cue for larval settlement is size (~25 mm), with size in part being a function of temperature - most common at 15-20°C [@steimle1999_scup]. Larval diets are poorly understood, but larvae have been successfully reared on zooplankton in lab experiments [@bowman1987food]. Early juveniles consume copepods, so larval scup may also consume copepods [@michelman1988biology]. Larval predators include plantivorous medusae, crustaceans, and fishes [@steimle1999_scup]. Juveniles have been caught in intertidal and subtidal habitats, over sand, silty-sand, shell, mud, mussel beds and eelgrass (*Zosteria marina*). Offshore winter-spring distributions indicate that they occur in habitats ranging from relatively flat, open, sandy-silty bottoms to the head of submarine canyons, and other areas with topographical relief and varying sediments. Juveniles can use biogenic depressions, sand wave troughs, and possibly mollusk shell fields for shelter in winter, suggesting structure may be important to juveniles [@able1998first]. Juveniles primarily feed on polychaetes, epibenthic amphipods and other small crustaceans, mollusks, and fish eggs and larvae [@steimle1999_scup]. Prey size increases with individual size, with copepods and mysids being important to early juveniles and bivalve mollusks consumed by larger fish [@bowman1987food]. Adult scup are also benthic feeders and forage on a variety of prey, including small crustaceans (including zooplankton), polychaetes, mollusks, small squid, vegetable detritus, insect larvae, hydroids, sand dollars, amphipods, and small fish [@steimle1999_scup]. Similar to juveniles, their prey size scales with individual size. Diets can vary based on location, and season. Feeding slows or stops entirely in winter [@byron2010stability]. Adults are primarily benthic and utilize both estuarine and shelf benthos [@steimle1999_scup]. Age at maturity is around 2 years of age, with a maximum age between 15-20 years [@nefsc2015sixtieth]. Scup undertake extensive inshore/offshore migrations driven by temperature preferences. They migrate south and offshore in the fall and return to inshore bays and estuaries in the spring to spawn. Scup can become inactive or dormant if caught in cold pockets < 6°C during their inshore spring migration and prefer temperatures 9-24°C [@steimle1999_scup]. The North Atlantic scup fishery is managed jointly by the Mid-Atlantic Fishery Management Council and Atlantic States Marine Fisheries Commission. The 2025 Management Track Assessment reported the Atlantic coast stock as not overfished and overfishing not occurring [@nefsc2025scup]. 

## Figures

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Scup_exposure_allvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Scup_exposure_impvars.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Scup_distribution_center_of_gravity.png")
```

```{r,echo=FALSE, fig.align='left'}
here::here("test_images/Scup_distribution_area_of_high_probability.png")
```

```{r,echo=FALSE, out.width="200%", fig.align='left'}
here::here("test_images/Scup_sdm.png")
```

## References
